Temporality of the Victorian Hospital Patient Transfer Network

### 2025-06-03 reports/paper.qmd
Upd rep

---
title: "Temporality of the Victorian Hospital Patient Transfer Network"
format:
  plos-pdf:
    number-sections: false
    journal:
      id: ploscompbio
    include-in-header:
      text: |
        \usepackage{setspace}
        \doublespacing
    keep-tex: true
author:
  - name: Michael J Lydeamore
    affiliations:
      - ref: ebs
      - ref: vicdh
    email: michael.lydeamore@monash.edu
    corresponding: true
  - name: David Wu
    affiliations:
      - ref: ebs
  - name: Tjibbe Donker
    affiliations:
      - ref: freiburg
  - name: Ben Cooper
    affiliations:
      - ref: oxford
  - name: Marion Easton
    affiliations:
      - ref: vicdh
  - name: Nicholas Geard
    affiliations:
      - ref: uom
  - name: Claire Gorrie
    affiliations:
      - ref: doherty
      - ref: mdu
  - name: Daneeta Hennessy
    affiliations:
      - ref: vicdh
  - name: Benjamin Howden
    affiliations:
      - ref: mdu
  - name: Anton Y Peleg
    affiliations:
      - ref: id
  - name: Charlie Higgs
    affiliations:
      - ref: mdu
  - name: Andrew Wilson
    affiliations:
      - ref: vicdh
  - name: Andrew J Stewardson
    affiliations:
      - ref: id

@@

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from matplotlib import pyplot as plt, dates as mdates
import seaborn as sns
import calendar
from itertools import accumulate
import datetime

@@

sns.set_theme(
  context='paper',
  style='white',
  palette='colorblind',
  font='cmr10',
)
```

@@

For the temporal network, we make a choice of the time discretisation $\omega$. That is, events that occur within a time window $[\tau, \tau+\omega)$ are collapsed and aggregated to occur at time $\tau$. Over the observation (and simulation) period, we will have a series of time window boundaries $\{\tau_0, \tau_1, \dots\}$. We denote the smallest value in that set larger than a given time $t$ to be $\lceil{}t\rceil{}_\omega$
For the readmission process, we assume that individuals that would readmit within a given time window $[\tau, \tau+\omega)$ do so uniformly. 
This yields:

@@

fig = plt.figure()
ax = fig.add_subplot()
for model, dist in hitting_time_dists.items():
    ax.plot(dist, np.linspace(0, 1, len(dist), endpoint=False), label=model)

@@

model_order = sorted(extent_dists.keys())
for ax, model in zip(axs, model_order):
  ax.hist(
    extent_dists[model],
    bins=extent_bins,
    density=True,
  )
  ax.set_ylabel(model)

axs[-1].set_xlabel('Extent')
```

@@

@@

>>>>

---
title: "Temporality of the Victorian Hospital Patient Transfer Network"
format:
  plos-pdf:
    number-sections: false
    journal:
      id: ploscompbio
    include-in-header:
      text: |
        \usepackage{setspace}
        \doublespacing
    keep-tex: true
    fig-format: png
    fig-dpi: 300
author:
  - name: Michael J Lydeamore
    affiliations:
      - ref: ebs
      - ref: vicdh
    email: michael.lydeamore@monash.edu
    corresponding: true
  - name: David Wu
    affiliations:
      - ref: ebs
  - name: Tjibbe Donker
    affiliations:
      - ref: freiburg
  - name: Ben Cooper
    affiliations:
      - ref: oxford
  - name: Marion Easton
    affiliations:
      - ref: vicdh
  - name: Nicholas Geard
    affiliations:
      - ref: uom
  - name: Claire Gorrie
    affiliations:
      - ref: doherty
      - ref: mdu
  - name: Daneeta Hennessy
    affiliations:
      - ref: vicdh
  - name: Benjamin Howden
    affiliations:
      - ref: mdu
  - name: Anton Y Peleg
    affiliations:
      - ref: id
  - name: Charlie Higgs
    affiliations:
      - ref: mdu
  - name: Andrew Wilson
    affiliations:
      - ref: vicdh
  - name: Andrew J Stewardson
    affiliations:
      - ref: id

@@

```{python}
#| label: compute-imports
import numpy as np
from pathlib import Path
import polars as pl
from matplotlib import pyplot as plt, dates as mdates, colors as mcolors
import seaborn as sns
import calendar
from itertools import accumulate
from collections import defaultdict
import datetime

@@

sns.set_theme(
  context='paper',
  style='ticks',
  palette='colorblind',
  font='serif',
)
```

@@

For the temporal network, we make a choice of the time discretisation $\omega$. Events that occur within a time window $[\tau, \tau+\omega)$ are collapsed and aggregated to occur at time $\tau$. Over the observation (and simulation) period, we will have a series of time window boundaries $\{\tau_0, \tau_1, \dots\}$. We denote the smallest value in that set larger than a given time $t$ to be $\lceil{}t\rceil{}_\omega$
For the readmission process, we assume that individuals that would readmit within a given time window $[\tau, \tau+\omega)$ do so uniformly. 
This yields:

@@

fig = plt.figure()
ax = fig.add_subplot()
model_order = ["naive_static", "static", "snapshot", "temporal"] 
for model in model_order:
    dist = hitting_time_dists[model]
    ax.plot(dist, np.linspace(0, 1, len(dist), endpoint=False), label=model.replace('_', ' '))

@@

model_order = ['temporal', 'snapshot', 'static', 'naive_static'] 
for ax, model in zip(axs, model_order):
  ax.hist(
    extent_dists[model],
    bins=extent_bins,
    density=True,
  )
  ax.set_ylabel(model.replace('_', '\n'))
  sns.despine(ax=ax)

_ = axs[-1].set_xlabel('Extent')
```

One striking feature

```{python}
histories = {
  dir.stem: np.loadtxt(dir/"hosp_presence.csv", delimiter=',')
  for dir in (root / "simulations/zero_sims_resized/" ).iterdir()
}
```
```{python}
#| label: fig-homeprop-heatmap
#| fig-cap: Distribution of the number of indivdiuals at home (not in the hospital system) over the duration of the simulation for each model type.

_BASE_DATE = np.datetime64('2011-01', 'M')
_MAX_DATE = np.datetime64('2018-12', 'M')
def set_xdatelabels(hist, ax, spacing=12, dt=1.0):
    # here, we want to get succinct date formats
    # we map the times onto the timedeltas of the simulations
    N = hist.shape[1]
    emp_max_date = (np.timedelta64(int(N*dt), 'D') + _BASE_DATE).astype('datetime64[M]')
    ax_max_date = np.min([emp_max_date, _MAX_DATE])

    dates = np.arange(_BASE_DATE, ax_max_date, np.timedelta64(spacing, 'M'))

    label_pos = (dates.astype('datetime64[D]').astype('int64') - _BASE_DATE.astype('datetime64[D]').astype('int64')) / dt

    ax.set_xticks(label_pos, dates.astype('datetime64[Y]'), rotation=45, rotation_mode='anchor')

    return ax.get_xticks()

_DTS = defaultdict(lambda: 1.0)
_DTS['temporal'] = 0.5
def single_heatmap(model, hist, bins=(540, 31), norm=None, ax=None, cbar=False):
    if ax is None:
        fig, ax = plt.subplots()

    if norm is None:
        norm = mcolors.Normalize()

    dt = _DTS[model]
    ts = (np.arange(hist.shape[1]) * dt).reshape((1, -1))
    ts_arr = np.repeat(ts, hist.shape[0], axis=0)

    *_, coll = ax.hist2d(
        ts_arr.flatten(), 
        30 - hist.flatten(),
        bins=bins,
        density=True,
        norm=norm,
    )

    if cbar:
        ax.figure.colorbar(coll, ax=ax)

    return coll, ax

_ax_order = {
    'temporal': 0,
    'snapshot': 1,
    'static': 2,
    'naive_static': 3,
}
def final_heatmap(hists, norm=mcolors.PowerNorm(0.5)):

    fig, axs = plt.subplots(nrows=4, figsize=[8.25, 5.25], sharex=True)
    axs_flat = axs.flatten()
    for k, h in hists.items():
        ax = axs_flat[_ax_order[k]]
        coll, _ = single_heatmap(k, h, norm=norm, ax=ax)
        norm = coll.norm
        ax.set_ylabel(k.replace('_', '\n'))
    set_xdatelabels(h, axs_flat[-1])
    fig.subplots_adjust(hspace=0.3)
    fig.colorbar(coll, ax=axs, label='Density', extend='max')
    fig.supylabel("Number at home ($N_0=30$)")

    return fig, axs

_ = final_heatmap(histories)
```

@@

# References
::: {#refs}
:::

::: {.content-visisble when-format="pdf"}
\newpage
:::

@@

## Uniform dissapation of at-home individuals

We know that for a random variable $t$ uniformly distributed between $[\tau, \tau+\omega]$, the pdf and survival functions are, respectively:

$$
\begin{gathered}
f(t) = \frac{1}{\omega}, \\
S(t) = \frac{(\tau+\omega) - t}{\omega},
\end{gathered}
$$

which results in a hazard function

$$
\lambda(t) = \frac{1}{(\tau + \omega) - t}.
$$

This hazard function can be simulated approximately in our DTMC framework by taking small time steps, and modelling the number of individuals $dN$ that move in a time step $dt$, where $n(t)$ is the number of inidividuals still present at time $t$ as

$$dN(t) \sim \text{Binomial}(n(t), dt\lambda(t)).$$
